---
title: Reading in Data 
--- 

Data files can come in multiple formats but some of the most common types are plain text files that have each column delimited by either tabs `"\t"` or commas `","`. R can also read in files that saved in excel format among many many others.   

## File Formats

### Plain Text Files
For reading in plain text files we will being using the package `readr`
```{r}
library(readr)
```

For reading in tab delimited files we have `read_tsv()` and for comma delimited files we have `read_csv()`
Here is a [link](../datasets/WorEpi.tab.txt) to a tab delimited file
Also a helpful function might be to use `file.choose()` 
```{r}
library(readr)
worEpi = read_tsv("datasets/WorEpi.tab.txt")

library(dplyr) # to use glimpse(), an improved version of str()
glimpse(worEpi)
```
or 
```{r, eval=FALSE}
library(readr)
library(dplyr)
worEpi = read_tsv(file.choose())
glimpse(worEpi)
head(worEpi)
```
Here is a [link](datasets/Mel.csv) to a comma separated file 
```{r}
library(readr)
melanoma = read_csv("datasets/Mel.csv")
glimpse(melanoma)
head(melanoma)
```
or 
```{r, eval=FALSE}
library(readr)
melanoma = read_csv(file.choose())
glimpse(melanoma)

```

## Rownames  
Sometimes your data will have row names as well, which the functions of readr will read in and put as the first column without a name and when using the readr functions it will be named "...1"     

```{r}
temps = read_tsv("datasets/Temperatures.txt")
print(temps)
```



### Excel Sheets
For excel sheets we have the library `readxl`
```{r, eval=T}
library(readxl)
```

Here is a [link](datasets/Both.xlsx) to an excel sheet of both previous files as separate sheets
```{r}
library(readxl)
melanoma = read_excel("datasets/Both.xlsx")
glimpse(melanoma)
head(melanoma)
```
By default it just reads the first sheet, you can tell it which sheet to read
```{r}
library(readxl)
worEpi = read_excel("datasets/Both.xlsx", sheet = 2)
glimpse(worEpi)
head(worEpi)
```

And to list all sheets available for reading 
```{r}
library(readxl)
sheets = excel_sheets("datasets/Both.xlsx")
print(sheets)
```

You can also read in by giving the sheet's name, to get what the sheets names are you can use 
```{r}
library(readxl)
print(excel_sheets("datasets/Both.xlsx"))

worEpi = read_excel("datasets/Both.xlsx", sheet = "Sheet2")
head(worEpi)

melanoma = read_excel("datasets/Both.xlsx", sheet = "Mel")
head(melanoma)

```


### SAS and SPSS
For reading data from SAS of SPSS files we have the library `haven`.  I don't have any example datasets but the functions are below

```{r, eval=FALSE}
library(haven)
read_sas()
read_spss()
read_stata()
```
       

